---
title: "Unsupervised classification (clustering) of satellite images"
author: "Krzysztof Dyba"
output:
  html_document:
    toc: yes
    toc_float: true
date: "29 August 2023"
---

## Data loading

```{r message=FALSE, warning=FALSE}
# load the package
library("terra")
```

In the first step, we need to create a list of files (rasters) that we are
going to load. To do this, we can use the `list.files()` function, which
takes a path to a folder with files as an argument. In addition, we must
indicate what kind of files we want to load (`pattern = "\\.TIF$"`) and
return full paths to the files (`full.names = TRUE`).

```{r}
# list files from a directory
files = list.files("dane_cut/", pattern = "\\.TIF$", full.names = TRUE)
files
```
 
Once we have created a list of files, we can load them using the `rast()`
function from the **terra** package and then display the metadata.

```{r}
# load raster data
landsat = rast(files)
landsat # calling the object displays the metadata
```

We can also shorten or rename the spectral bands. Before this, make sure
that the bands are loaded in the correct order.

```{r}
names(landsat) # original names
names(landsat) = paste0("B", 1:7) # shorten the names
names(landsat) # new names

# or alternatively rename
# names(landsat) = c("Ultra Blue", "Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2")
```

Loading vector data is done in an analogous way using the `vect()` function.

```{r}
# load vector data
poly = vect("Poznan.gpkg")
poly
```

As we can see from the metadata, raster and vector data have different
coordinate reference systems (CRS), which is troublesome. The easiest way
is to transform the vector data into a raster's CRS and we can do it with
the `project()` function and specifying the EPSG code.

```{r}
poly = project(poly, "EPSG:32633")
crs(poly, describe = TRUE)$code # check EPSG after transformation
```

Now we can prepare a simple visualization using the near infrared band (NIR; B5)
and polygon as an example.

```{r}
# visualization
plot(landsat[[5]], main = "Poznan commune") # alternatively: plot(landsat[["B5"]])
plot(poly, add = TRUE)
```

## Raster processing

The extent of our analysis area is limited to the Poznan commune, while the
satellite scene has a much larger extent. In such a situation, we can crop the
rasters, so that their further processing will be faster and the results
will take up less space on disk. The `crop()` function is used to crop the
rasters, and we need to specify the raster and vector as arguments.

Note that the rasters are represented as matrices, so the cropping is done
to the bounding box. To include only the area of our polygon in the analysis,
we need to mask the pixels outside the boundary. This can be done by setting
the `mask = TRUE` argument in the aforementioned function.

```{r}
landsat = crop(landsat, poly, mask = TRUE)
plot(landsat[[5]], main = "Poznan commune")
plot(poly, add = TRUE)
```

In the next step, we can easily check the descriptive statistics of our dataset.

```{r warning=FALSE}
summary(landsat)
```

As we can see, the spectral reflectance values are several thousand for each
band and are encoded as integers. The spectral reflectance should be in the
range from 0 to 1. Therefore, we need to scale our data using the following
equation (this only applies to reflectance, not temperature!):

$$x = x \cdot 0.0000275 - 0.2$$

For example, the pixel value in the near infrared (NIR) band is 15000. Using the
above formula, we need to multiply this value by 0.0000275 (scale factor) and
then subtract 0.2 (offset). As a result, we will get a reflection equal
to 0.2125. Note that each product/collection has a different formula and
it is necessary to consult the documentation.

We don't need to apply this formula separately for each band in the loop
because the arithmetic operations in the **terra** package is applied to
all bands by default.

```{r warning=FALSE}
landsat = landsat * 2.75e-05 - 0.2
summary(landsat)
```

We can still see that some values exceed the range of 0 to 1. These are
outliers that are usually associated with incorrect measurement or
oversaturation. We can solve this problem in two ways:

1. Replace these values with missing data (*NA*).
2. Trim to minimum and maximum value.

The first way can cause us to lose a large part of the dataset. The second way,
on the other hand, can cause skewed results.

```{r}
# method 1
landsat[landsat < 0] = NA
landsat[landsat > 1] = NA
```

```{r eval=FALSE}
# method 2
landsat[landsat < 0] = 0
landsat[landsat > 1] = 1
```

After scaling the values, we can display the RGB composition. In this case,
instead of `plot()` function, use `plotRGB()` function and define the order of
red, green and blue bands. In addition, we need to specify the maximum
reflection value for the bands (in our case `scale = 1`). It often happens
that compositions are too dark/bright, then is helpful to apply color stretching
using the `stretch = "lin"` or `stretch = "hist"` argument.

```{r}
# plotRGB(landsat, r = 4, g = 3, b = 2, scale = 1)
plotRGB(landsat, r = 4, g = 3, b = 2, scale = 1, stretch = "lin")
```

## Clustering

```{r}
# load clustering package
library("cluster")
```

Data for modeling must be prepared in an appropriate way. Classification
models most often require a matrix or data frame at the training stage.
Raster data can be transformed into a matrix using the `values()` function.

```{r}
mat = values(landsat)
nrow(mat) # print number of rows/pixels
```

Using the interactive `View()` function, we can see what our matrix looks like.

```{r, eval=FALSE}
View(mat)
```

As we can see, a lot of its values are marked as missing data (mostly they are
outside the analysis area). Usually models do not support *NA*, so we need to
remove them. A dedicated function `na.omit()` can be used for this.

```{r}
mat_omit = na.omit(mat)
nrow(mat_omit)
```

Now we will move on to the next stage of the analysis, which is to
train the model. There are many clustering methods and models
(see [CRAN Task View](https://cran.r-project.org/web/views/Cluster.html)),
but in this example we will use the simplest
[k-means](https://en.wikipedia.org/wiki/K-means_clustering) clustering method.
This model only requires the number of groups/clusters (`centers` argument) to
be given in advance. This is a stochastic algorithm, so it returns different
results each time. To make the analysis repeatable we need to set the seed of
randomness -- `set.seed()`.

```{r warning=FALSE}
set.seed(1)
mdl = kmeans(mat_omit, centers = 6)
```

As a result of the above operation, we received, among others:

1. Calculated group averages for each band `(mdl$centers)`.
2. Vector with classified matrix (pixels) values `(mdl$cluster)`.

Let's display these objects:

```{r}
mdl$centers
```

```{r}
head(mdl$cluster) # display the first 6 elements
```

This means that the first row (representing a single pixel) belongs to group 2,
the second to group 2, the third to group 2, and so on.

## Validation

An inseparable element of modeling is the validation of the created models.
The challenge is to choose the right grouping method for a specific dataset
and determine the appropriate number of groups. Note that increasing the number
of clusters increases the similarity between objects in a cluster, but with
more clusters, their interpretation becomes more difficult.

The most common way to validate clustering results is to use internal metrics,
such as [Dunn index](https://en.wikipedia.org/wiki/Dunn_index),
[Davies–Bouldin index](https://en.wikipedia.org/wiki/Davies%E2%80%93Bouldin_index)
or [silhouette index](https://en.wikipedia.org/wiki/Silhouette_(clustering)).
In this example, we will use the latter.

Silhouette index evaluates the compactness and separation of the clusters
based on the distances between objects within the same cluster and between
objects in different clusters. The values of this index range from -1 to 1.
A value close to 1 indicates that the object is well clustered and is far away
from neighboring clusters. A value close to -1 suggests that the object may have
been assigned to the wrong cluster. A value near 0 indicates that the object
is very close to the boundary between different clusters. Thus, in general,
a higher value of this index indicates better clustering results. See the
`cluster::silhouette()` documentation for more details.

Now let's try to calculate this index. Basically, using this index requires
calculating the similarity of each object with every object, which in our case
is an impossible task (our dataset consists of 2.5 million objects). To do this,
we need to draw a smaller sample (let's say $n = 10 000$). In the function,
we need to specify two objects -- a vector with clusters and a dissimilarity
matrix, which can be calculated in advance using the `dist()` function. 

```{r}
set.seed(1)
# draw sample indexes
idx = sample(1:nrow(mat_omit), size = 10000)
head(idx)
```

```{r}
# calculate silhouette index
sil = silhouette(mdl$cluster[idx], dist(mat_omit[idx, ]))
summary(sil)
```

The average cluster convergence is `r round(mean(sil[, 3]), 2)`. This is not the
best result. We should try to increase the number of clusters or use another
clustering method. We can also check the results on the chart.

```{r}
colors = rainbow(n = 6) # define the color scheme
plot(sil, border = NA, col = colors, main = "Silhouette Index")
```

## Interpretation

The essence of grouping is to create a group of similar objects, but our task
is to interpret what the created groups represent and name them. Interpretation
is a difficult task, and often the results are unclear. For this purpose, it
is necessary to analyze the descriptive statistics of groups and to use various
plots and map compositions. Knowledge of the spectral properties of objects is
also very useful.

So let's try to interpret the obtained clusters by assisting with a boxplot.
The **ggplot2** package works best for creating visualizations. Here you
can find a free [manual](https://ggplot2-book.org/) and ready-made
[recipes](https://r-graphics.org/).

**ggplot2** requires preparing the dataset to the appropriate form. The data
must be presented as a data frame in the so-called long form (multiple rows),
while base functions for visualization require wide form (multiple columns).
This change can be made easily using the **tidyr** package.

```{r message=FALSE}
# install.packages(c("tidyr", "ggplot2"))
library("tidyr") # data transformation
library("ggplot2") # data visualization
```

As we noted earlier, our dataset is quite large and there is no need to present
all the data. We can do it more efficiently by using a previously drawn sample.
Now we need to combine the drawn rows from the matrix with the corresponding
clusters (`cbind()`). Then we will convert the matrix into a data frame
(`as.data.frame()`).

```{r}
stats = cbind(mat_omit[idx, ], cluster = mdl$cluster[idx])
stats = as.data.frame(stats)
head(stats)
```

The displayed data is in a wide form (each spectral band is stored in a separate
column). Now we need to change the form, in which we get two columns -- the band
and the value. To do this, we will use the `pivot_longer()` function.

```{r}
stats = pivot_longer(stats, cols = 1:7, names_to = "band", values_to = "value")
# change the data type to factor
stats$cluster = as.factor(stats$cluster)
stats$band = as.factor(stats$band)
head(stats)
```

The data structure is already prepared. Now let's create a simple boxplot.

```{r}
ggplot(stats, aes(x = band, y = value, fill = cluster)) +
  geom_boxplot()
```

By changing a few default parameters, we can improve the perception of this figure.

```{r}
ggplot(stats, aes(x = band, y = value, fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = colors) +
  facet_wrap(vars(cluster)) +
  xlab("Spectral band") +
  ylab("Reflectance") +
  theme_light()
```

NAPISAC INTERPRETACJE


## Final map

```{r}
vec = rep(NA, ncell(landsat)) # prepare an empty vector
vec[complete.cases(mat)] = mdl$cluster # assign clusters
clustering = rast(landsat, nlyrs = 1, vals = vec) # create raster
plot(clustering, type = "classes", col = colors)
```

If the result is satisfactory, we can save it using the `writeRaster()` function.
Such a file can be later loaded in **R** or another program that supports spatial
data (e.g. **QGIS**). In addition, for categorical data, it is useful to set the
datatype as `Byte` / `INT1U` when saving.

```{r eval=FALSE}
writeRaster(clustering, "clustering.tif", datatype = "INT1U")
```

## Summary






